---
title: "Exploring Environmental Justice in LA"
author: Henry Oliver
format: html
warning: FALSE
message: FALSE
---

#### Background

Present-day environmental justice may reflect legacies of injustice in the past. The United States has a long history of racial segregation which is still visible. During the 1930’s the Home Owners’ Loan Corporation (HOLC), as part of the New Deal, rated neighborhoods based on their perceived safety for real estate investment. Their ranking system, (A (green), B (blue), C (yellow), D (red)) was then used to block access to loans for home ownership. Colloquially known as “redlining”, this practice has had widely-documented consequences not only for community wealth, but also health (1). Redlined neighborhoods have less greenery (2) and are hotter than other neighborhoods (3).

A recent study found that redlining has not only affected the environments communities are exposed to, it has also shaped our observations of biodiversity.4 Community or citizen science, whereby individuals share observations of species, is generating an enormous volume of data. Ellis-Soto and co-authors found that redlined neighborhoods remain the most undersampled areas across 195 US cities. This gap is highly concerning, because conservation decisions are made based on these data.


## 1) Prepare environment
```{r}
# Load library
library(tidyverse)
library(here)
library(tmap)
library(stars)
library(sf)
library(spData)
library(janitor)
library(knitr)
library(ggpubr)
library(ggthemes)


# Read in data
# EJScreen - Census
ejscreen <- st_read(here::here('data/ejscreen/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb')) %>% 
  clean_names() %>% 
  # Filter to LA County
  filter((st_abbrev == 'CA') & (cnty_name == 'Los Angeles County')) %>%
  select(shape_length, shape_area, Shape, id,
st_abbrev, cnty_name, region, acstotpop, lowincome, p_pm25, p_lifeexppct, p_lowincpct)
  
# HOLC Redlining
holc <- st_read(here::here('data/mapping-inequality/mapping-inequality-los-angeles.json')) %>% 
  st_transform(crs = st_crs(ejscreen)) %>% # Transform CRS
  clean_names()
  
# Biodiversity observations
birds <- st_read(here::here('data/gbif-birds-LA')) %>% 
  st_transform(crs = st_crs(ejscreen)) %>%  # Transform CRS
  clean_names()

  
```


## 2) Historical Redlining Neighborhoods Map
```{r}
map1 <-
  tm_basemap(server = "https://tile.openstreetmap.org/{z}/{x}/{y}.png", alpha = 0.7) +
  tm_shape(holc) +
    tm_polygons(fill = 'grade',
                palette = c(
                  "A" = "lightgreen",
                  "B" = "lightblue",
                  "C" = "#FFFF66",
                  "D" = "#FA2A55")) +
  tm_title(text = "HOLC Grades in LA County",
            frame = FALSE) +
  tm_scale_bar(position = c(0,0.1)) +
  tm_compass()

map1
  

```

3) HOLC Grade Summary Table

```{r}
# Join ejscreen and holc by Shape (ejscreen) and geometry (holc)
joined = st_join(ejscreen, holc) %>%
  st_drop_geometry()

# Get the count of each grade / total census blocks
pct_A <- (sum(joined$grade == 'A', na.rm = TRUE) / nrow(joined)) * 100
pct_B <- (sum(joined$grade == 'B', na.rm = TRUE) / nrow(joined)) * 100
pct_C <- (sum(joined$grade == 'C', na.rm = TRUE) / nrow(joined)) * 100
pct_D <- (sum(joined$grade == 'D', na.rm = TRUE) / nrow(joined)) * 100
pct_na <- (sum(is.na(joined$grade)) / nrow(joined)) * 100

counts <- joined %>% 
  group_by(grade) %>% 
  summarize(tbl_counts = sum(lowincome))

pct_df <- data.frame(grade = c('A', 'B', 'C', 'D', 'No Grade'),
                     count = counts$tbl_counts,
                     percentage = c(pct_A, pct_B, pct_C, pct_D, pct_na))

pct_table <- kable(pct_df, col.names = c("Grade", "Count", "Percentage"), caption = "HOLC Grade Percentages in LA County")

pct_table
```



4) Summary of low income, air quality, and low life expectancy by HOLC Grade

```{r}
# Low income housing
low_income <- joined %>% 
  group_by(grade) %>% #Group by grade
  summarize(low_inc_mean = mean(p_lowincpct)) #Get the mean of percentile low income per grade
   
  
plot1 <- ggplot(low_income, aes(x = grade, y = low_inc_mean)) +
  geom_bar(stat = 'identity', fill = '#CE5425') + #stat = 'identity' tells ggplot that the y values are actual data values, not just counts
  labs(title = "Low Income",
       x = "HOLC Grade",
       y = "Mean Percentile Low Income",
       caption = "Fig. 1") +
  theme_clean() +
  theme(plot.caption = element_text(hjust = 0)) # Move caption to the left

#Air quality (pm25)
pct_pm25 <- joined %>% 
  group_by(grade) %>% #Group by grade
  summarize(pm25 = mean(p_pm25, na.rm=TRUE))

plot2 <- ggplot(pct_pm25, aes(x = grade, y = pm25)) +
  geom_bar(stat = 'identity', fill = '#5E899E') +
  labs(title = "PM25", x = "HOLC Grade",
       y = "Mean Percentile PM25",
       caption = "Fig. 2") +
  theme_clean() +
  theme(plot.caption = element_text(hjust = 0))
 
# Life expectancy
life_exp <- joined %>% 
  group_by(grade) %>% 
  summarize(lle = mean(p_lifeexppct, na.rm=TRUE))

plot3 <- ggplot(life_exp, aes(x = grade, y = lle)) +
  geom_bar(stat = 'identity', fill = '#4b7a47') +
  labs(title = "Low Life Expectancy", x = "HOLC Grade",
       y = "Mean Percentile Low Life Expectancy",
       caption = "Fig. 3") +
  theme_clean() +
  theme(plot.caption = element_text(hjust = 0))


annotate_figure( # Create caption-able object
  ggarrange(plot1, plot2, plot3, # Combine plots
                          ncol =3, nrow = 1, # Specify shape
                          widths = c(1, 1, 1.2)), # Specify individual widths
                          # Caption top
          top = text_grob('Environmental Justice by HOLC Grades, 2021-2023')) 



```

## 5) Interpretation

  The figures above display analysis for three key metrics of quality of life as they correspond to redlining districts in LA County. Fig. 1 displays the mean percentile of low income homes found in each LA County census block. "Low income" is defined to be homes which have income equal to or less than twice the federal poverty level (USEPA). Fig. 2 displays the mean percentile of homes per census block that were found to have airborne particles of size 2.5 micrometers or smaller (PM25), a direct indicator of air quality. Fig. 3 displays the mean percentile of individuals with lower than average national life expectancy per census block. For all three metrics, we see that lower HOLC grades ('A' being the highest, and 'D' being the lowest) are associated with higher poverty, higher air pollution, and higher mortality. This pattern supports the argument that when the Home Owners' Loan Corporation (HOLC) classified neighborhoods by investment risk in the 1930s, it effectively established a housing value hierarchy -- boosting property values in higher quality-of-life areas while devaluing those in lower quality-of-life, often low-income, communities. The fact that these patterns are easily visible nearly a century after redlining took place shows the long-lasting and still-present impact of classist policies in the early 20th century.


## 6) Biodiversity and Redlining

1) Create a figure summarizing the percent of observations within redlined neighborhood within each HOLC grade.
```{r}
# Check geometry type for birds
print(paste("Geometry type for birds is", unique(st_geometry_type(birds))))

# Confirm CRS matches
if (st_crs(holc) == st_crs(birds)) {
  print("CRS for birds and holc are matching")
} else {
  print("CRS DOES NOT MATCH")
}

# Join birds and holc data
birds_holc <- st_join(birds, holc, join = st_within) # Using st_within because the birds are points, so they cannot be partially in/out of HOLC geometries

# Get counts of bird observations per HOLC grade
bird_grades <- birds_holc %>% 
  st_drop_geometry() %>% #Drop geometries
  filter(!is.na(grade)) %>% # Drop NAs because there are so many it will get in the way of visualizing our analysis
  count(grade, name = "count_grade") %>% # Return count by grade, rename this 'count'
  mutate(percentage = (count_grade / sum(count_grade)) * 100)


fig4 <- ggplot(bird_grades, aes(x = grade, y = percentage)) +
  geom_bar(stat = 'identity', fill = "#FA8072") +
  labs(title = "Percentage of Bird Observations by HOLC Grade",
       x = "HOLC Grade",
       y = "Percentage of Bird Observations",
       caption = "Fig. 4") +
  theme_clean()  +
  theme(plot.caption = element_text(hjust = 0)) #Position caption on left
  
fig4
```


## References

https://19january2017snapshot.epa.gov/ejscreen/overview-demographic-indicators-ejscreen_.html


